A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-12 tidyindex.qmd
Update conclusions

The paper introduces a data pipeline for constructing and analyzing indexes within the tidy framework. The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.

One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity analysis. While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives. The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains. These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).

>>>>

The paper introduces a novel data pipeline for constructing and analyzing indexes within the tidy framework. The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.

One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity. The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains. These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility. This work highlights the importance of carefully constructing statistical indexes and introduces novel computational methods for studying and calculating statistical indexes and their uncertainties.

### 2024-04-12 tidyindex.tex
Update conclusions

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

The paper introduces a data pipeline for constructing and analyzing
indexes within the tidy framework. The pipeline comprises nine modules
including temporal and spatial aggregation, variable transformation and
combination, distribution fitting, benchmark setting, and index
communication. The proposed pipeline is implemented as an R package,
\texttt{tidyindex}, available on CRAN.

One of the contributions of our work is its emphasis on aspects often
overlooked in index construction: uncertainty quantification and
sensitivity analysis. While indexes are widely constructed and used to
inform decision-making, such as on resource allocation, little effort is
made to justify choices made during indexes construction and to
experiment with alternatives. The pipeline proposed in the paper allows
for confidence interval calculation and sensitivity analysis to be
incorporated in the index construction process, for indexes across
various domains. These is illustrated through the three examples on
drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

The paper introduces a novel data pipeline for constructing and
analyzing indexes within the tidy framework. The pipeline comprises nine
modules including temporal and spatial aggregation, variable
transformation and combination, distribution fitting, benchmark setting,
and index communication. The proposed pipeline is implemented as an R
package, \texttt{tidyindex}, available on CRAN.

One of the contributions of our work is its emphasis on aspects often
overlooked in index construction: uncertainty quantification and
sensitivity. The pipeline proposed in the paper allows for confidence
interval calculation and sensitivity analysis to be incorporated in the
index construction process, for indexes across various domains. These is
illustrated through the three examples on drought indexes (SPI and SPEI)
and the Global Gender Gap Index (GGGI). While indexes are widely
constructed and used to inform decision-making, such as in resource
allocation and policy making worldwide, little effort is made to justify
the choices made during index construction or to experiment with
alternatives while providing means for reproducibility. This work
highlights the importance of carefully constructing statistical indexes
and introduces novel computational methods for studying and calculating
statistical indexes and their uncertainties.
